The Abundance of Bullet Groups in $\Lambda$CDM

### 2014-04-16 paper/ms.tex
updated

\title{The Abundance of Bullet-like Groups in $\Lambda$CDM}
\author{J. G. Fern\'andez-Trincado$^{1,2,3}$, J. E. Forero-Romero$^1$,
  G.Foex$^4$, T. Verdugo$^3$ and V. Motta$^4$} 
\affil{$^1$ Departamento de F\'{i}sica, Universidad de los Andes,
  Cra. 1 No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\ 
       $^2$ Institute Utinam, CNRS UMR6213, Universit\'e de
  Franche-Comt\'e, OSU THETA de Franche-Comt\'e-Bourgogne,
  Besan\c{c}on, France\\ 
       $^3$ Centro de Investigaciones de Astronom\'ia, AP 264,
  M\'erida 5101-A, Venezuela\\
  $^4$ Instituto de F\'isica y Astronom\'ia, Universidad de
  Valpara\'iso, Avda. Gran Breta\~na 1111, Playa Ancha, Valpara\'iso
  2360102, Chile  
}
\email{jfernandezt87@gmail.com}
\email{je.forero@uniandes.edu.co}
\begin{abstract}

@@

The Bullet Cluster (\bulla) provided a new kind of observational
evidence for the existence of dark matter
\citep{Markevitch2004,Clowe2006}. Since then it has been used to test
the Cold Dark Matter (CDM) paradigm itself by quantifying different
aspects such as the expected displacement between the dominant dark
matter and baryonic component in a $\Lambda$CDM Universe
\citep{ForeroRomero2010}, the substructure velocity required to
produce such displacement
\citep{Milosavljevic2007,Springel2007,Mastropietro20008} and its
abundance in large N-body cosmological simulations
\citep{Hayashi2006,Lee2012, Thompson2012}. It has also been used to
constrain the dark matter particle self-interaction cross section and
to explore possible extensions to the concordance cosmological model
\citep{Farrar2007,Lee2010}.

@@

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250\hMpc comoving on a side \citep{2011ApJ...740..120K}. The
simulation uses the ART code \citep{Kravtsov1997} to follow the
evolution of a dark matter density field from $z=80$ to $z=0$ sampled
with $2048^3$  particles. The cosmology used  corresponds to  the
spatially flat concordance model with the following parameters:  the
density parameter for matter (dark matter and baryons)
$\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the slope of
the primordial power spectrum $n=0.95$ and the amplitude of mass
density fluctuations (at redshift z$=$0) $\sigma_8=0.82$.  These
cosmological parameters are consistent with the nine-year Wilikinson
Microwave Anisotropy Probe (WMAP) results \citep{hinshaw_etal13}. A
detailed presentation of the simulation can be found in
\citet{2011ApJ...740..102K}.

@@

Each one of the pairs (host \& sub-halo) is considered as a
potential bullet system and is kept for the analysis described in the
next Section  by using the sub-halos as the tracer of the sub-dominant
dark matter clump in the merging cluster, i.e. the bullets.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_2.pdf}
\end{center}
\caption{2D histograms in the plane $\mu$-$D_{\rm off}$. The left
  panel corresponds to clusters and the right panel to groups. The data
  used to construct this histogram includes halos at all
  redshifts. This approximation to increase the statistical power of
  the sample is justified because the redshift evolution of this plot
  is weak.} 
\label{fig:geometry}
\end{figure*}

@@

For pairs on radial trajectories there are three regions of interest
in this plane that correspond to the three merging stages described
at the end of Section \ref{sec:geometry} assuming that the sub-halo
merges (or falls below the BDM detection threshold) right at its
second pass through the center of the host halo \citep{Poole2006}.

The first region has $\mu\sim-1$ and $D_{\rm off}>0.6$, which
locates the systems where a head-on collision has just started. The
second region has $\mu\sim 1$ and $D_{\rm off}<0.6$; at this stage
the collision continues after the first crossing of the host's
center. The low number of halos with radial infalling velocities and
displacements $D_{\rm off}>0.6$ suggests that this the maximum range of
radii for the apogee.  The third region corresponds to $\mu\sim-1$
and $D_{\rm off}<0.6$ that is the the secondary infall after apogee.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_3.pdf}
\end{center}
\caption{Integrated probability distribution for the estimated
  DM-gas displacements in the group and cluster samples. Continous (dashed)
  lines correspond to clusters (groups). The continuous black line
  marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and gas. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.} 
\label{fig:baryonic_displacements}
\end{figure*}

@@

J.E.F-R acknowledges support from Vicerrector\'ia de
Investigaciones through a FAPA starting grant.

>>>>

\title{The Abundance of Bullet Groups in $\Lambda$CDM}
\author{J. G. Fern\'andez-Trincado$^{1,2,3}$, J. E. Forero-Romero$^1$,
  G.Foex$^4$, T. Verdugo$^3$ and V. Motta$^4$} 
\affil{$^1$ Departamento de F\'{i}sica, Universidad de los Andes,
  Cra. 1 No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\ 
       $^2$ Institute Utinam, CNRS UMR6213, Universit\'e de
  Franche-Comt\'e, OSU THETA de Franche-Comt\'e-Bourgogne,
  Besan\c{c}on, France\\ 
       $^3$ Centro de Investigaciones de Astronom\'ia, AP 264,
  M\'erida 5101-A, Venezuela\\
  $^4$ Instituto de F\'isica y Astronom\'ia, Universidad de
  Valpara\'iso, Avda. Gran Breta\~na 1111, Playa Ancha, Valpara\'iso
  2360102, Chile  
}
\email{jfernandezt87@gmail.com}
\email{je.forero@uniandes.edu.co}
\begin{abstract}

@@

The Bullet Cluster (\bulla) provided a new kind of observational
evidence for the existence of dark matter
\citep{Markevitch2004,Clowe2006}. Since then it has been used to test
the Cold Dark Matter (CDM) paradigm itself by quantifying different
aspects such as the expected displacement between the dominant dark
matter and baryonic component in a $\Lambda$CDM Universe
\citep{ForeroRomero2010}, the substructure velocity required to
produce such displacement
\citep{Milosavljevic2007,Springel2007,Mastropietro2008} and its
abundance in large N-body cosmological simulations
\citep{Hayashi2006,Lee2012, Thompson2012}. It has also been used to
constrain the dark matter particle self-interaction cross section and
to explore possible extensions to the concordance cosmological model
\citep{Farrar2007,Lee2010}.

@@

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250\hMpc comoving on a side \citep{2011ApJ...740..102K}. The
simulation uses the ART code \citep{Kravtsov1997} to follow the
evolution of a dark matter density field from $z=80$ to $z=0$ sampled
with $2048^3$  particles. The cosmology used  corresponds to  the
spatially flat concordance model with the following parameters:  the
density parameter for matter (dark matter and baryons)
$\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the slope of
the primordial power spectrum $n=0.95$ and the amplitude of mass
density fluctuations (at redshift z$=$0) $\sigma_8=0.82$.  These
cosmological parameters are consistent with the nine-year Wilikinson
Microwave Anisotropy Probe (WMAP) results \citep{hinshaw_etal13}. A
detailed presentation of the simulation can be found in
\citet{2011ApJ...740..102K}.

@@

Each one of the pairs (host \& sub-halo) is considered as a
potential Bullet-like system and is kept for the analysis described in the
next Section  by using the sub-halos as the tracer of the sub-dominant
dark matter clump in the merging cluster, i.e. the bullets.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_2.pdf}
\end{center}
\caption{2D histograms in the plane $\mu$-$D_{\rm off}$ describing the
  geometry of the bullet encounter. The left
  panel corresponds to clusters and the right panel to groups. 
  The weak redshift evolution of the structure in these planes
  allows us to include halos at all redshifts to construct these figures.
  The three rectangles (continous, dashed, dashed-dotted)
  delimit stages of interest in a head-on collision as described in
  Section \ref{sec:geometry}.}
\label{fig:geometry}
\end{figure*}

@@

For pairs on radial trajectories there are three regions of interest
in this plane that correspond to the three merging stages described
at the end of Section \ref{sec:geometry} assuming that the sub-halo
merges (or falls below the BDM detection threshold) right at its
second pass through the center of the host halo \citep{Poole2006}.
These regions are shown as three different rectangled in Figure
\ref{fig:geometry}.

The first region has $\mu<-0.9$ and $D_{\rm off}>0.6$, which
locates the systems where a head-on collision has just started. The
second region has $\mu>0.9$ and $D_{\rm off}<0.6$; at this stage
the collision continues after the first crossing of the host's
center. The low number of halos with radial infalling velocities and
displacements $D_{\rm off}>0.6$ suggests that this the maximum range of
radii for the apogee.  The third region corresponds to $\mu<-0.9$
and $D_{\rm off}<0.6$ that is the the secondary infall after apogee.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.8\textwidth]{figure_3.pdf}
\end{center}
\caption{Integrated probability distribution for the estimated
  DM-gas displacements in the group and cluster samples. Continous (dashed)
  lines correspond to clusters (groups). The continuous black line
  marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and gas. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.} 
\label{fig:baryonic_displacements}
\end{figure*}

@@

We thank the referee for a report that improved the readibility of our
results. 
J.E.F-R acknowledges support from Vicerrector\'ia de
Investigaciones through a FAPA starting grant.
